---
title: '20200526-rethinking-chpt1_win10'
author: 'diaorch'
date: '5/26/2020'
output: 
    html_document: 
        toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# disable fancy quotes, 
# which is causing encoding errors in my Win10, R4.0, RStudio1.3.959, 
# as of 20200526
# see https://community.rstudio.com/t/unicode-replacement-character-issue-in-rstudio-only-within-r-markdown-files/7621/9
options(useFancyQuotes = FALSE)
```

## Installation of `rstan` on a Windows 10 machine

Below is a breakdown of steps needed to install the `rethinking` package with a freshly `R` and `RStudio` installation, starting from installing the dependencies. Codes are included below the step list.

1. install `Rtools`: download `Rtools` from and follow the instruction on the [CRAN "Using Rtools40 on Windows" page](https://cran.r-project.org/bin/windows/Rtools/), take notice to check if the version of `RTools` is compatible with *your* `R` version. Also, to "restart R" as instructed, assuming you are using `RStudio`, you would need to quit and reopen the RStudio, and only restarting the R session within RStudio would not update the path. 

2. install `devtools` if needed, you can check by running `find.package('devtools')`. If the package is not found (e.g. output `Error in find.package('devtools') : there is no package called ‘devtools’`), install by running `install.packages('devtools)`.

3. load `devtools` and verify `Rtools` installation by running `find_rtools()`. This step is ordered as such because the function `find_rtools()` is in the `devtools` package.

4. install `RStan` by running `install.packages("rstan", repos = "https://cloud.r-project.org/", dependencies = TRUE)`. This actually installs `RStan` from source, which is the reason for all previous steps needed in Windows.

5. install `rethinking` from source (GitHub) by running `devtools::install_github('rmcelreath/rethinking')`

```{r eval = FALSE}
# step 1
# download and install RTools40 from CRAN Rtools website, not shown
# put the location of the Rtools make utilities (bash, make, etc) on the PATH
# there are multiple ways to do it, one way is to run the next line in R
writeLines('PATH="${RTOOLS40_HOME}\\usr\\bin;${PATH}"', con = '~/.Renviron')
# restart R - restart the whole RStudio if in one, then verify the path
Sys.which('make')
## "C:\\rtools40\\usr\\bin\\make.exe"

# step 2
find.package('devtools')
## Error in find.package('devtools') : there is no package called ‘devtools’
if(!require('devtools')){
    install.packages('devtools', dependencies = TRUE)
}
## 20200526, Win10, the package `backports` needed to be explicitly installed 
## for `devtools` to be correctly loaded
## I suspect that is a compatibility issue in `backports` with R 4.0
## when prompted to choose whether to install from source instead of binary, 
## confirm to use the source (the source is of later versioni than the binary)
# install.packages('backports')

# step 3
library('devtools')
find_rtools()

# step 4
install.packages('rstan', 
                 repos = 'https://cloud.r-project.org/', dependencies = TRUE)
library('rstan')

# step 5
devtools::install_github('rmcelreath/rethinking')
library(rethinking)
help(package=rethinking)
## verify successful installation if in-system help opens up
```

## Linear regression with two independent categorical variables - an example with `stats::lm()`, `brms`, and `rethinking`

Consider a linear regression model in the form of: 

$$
y_i = \beta_0 + \beta_1 x_{1,i} + \beta_2 x_{2,i} + \beta_3 x_{1,i} x_{2,i} + \epsilon_{i}
$$

### Data simulation

What information we need for each sample as a starting point is:

1. "treated/untreated" by drug A: as a integer, in this case binary (0 or 1) 
2. "treated/untreated" by drug B: as a integer, in this case binary (0 or 1) 
3. "measurement of treatment effects": as a float, in this case randomly generated

``` {r}
set.seed(13)
drugA <- c(rep(0, 10), rep(1, 10))
drugB <- rep(c(rep(0, 5), rep(1, 5)), 2)
effect <- c(rnorm(5), rnorm(5) + 2, rnorm(5) + 3, rnorm(5) + 10)
alldat <- data.frame(
    'drugA' = drugA, 'drugB' = drugB, 'effect' = effect
)
print(alldat)
```

### Linear regression with two independent categorical variables - with `stats::lm()`

```{r}
twovarMod <- lm(effect ~ drugA + drugB + drugA:drugB, data = alldat)
print(summary(twovarMod))
```

### Linear regression with two independent categorical variables - with `brms`

Fitting a Bayesian Generalized Linear Multivariate Model:

(Expect a large chunk of output as the BRMS fit finishes up.)

```{r}
library('brms')
alldatFactor <- data.frame(
    'drugA' = factor(drugA, levels = c(0, 1)), 'drugB' = factor(drugB, levels = c(0, 1)), 'effect' = effect
)
brmCatMod <- brm(effect ~ drugA + drugB + drugA:drugB, data = alldatFactor)
```

As an interesting side note: it took me a good while for the BRMS fitting run to get pass compiling the C++ model. (There is always a balance between the result performance and the speed.)

```{r}
print(summary(brmCatMod))
conditional_effects(brmCatMod, effects = 'drugA:drugB')
# to see all effects (both individual and two-ways)
# remove the argument effects from the command above
```

`R` is always interesting when it comes to factors. What if we don't convert the `drugA` and `drugB` column from numerics to factors? As you can see in the `lm()` it does not cause *major* hiccup but in `brms::brm()`, if we do the following instead:

```{r message=FALSE, warning=FALSE}
brmMod <- brm(effect ~ drugA + drugB + drugA:drugB, data = alldat, silent = TRUE, refresh = 0)
conditional_effects(brmMod)
```

Notice the types of figures compared to the previous results with categorical inputs.

If we check the chains:

```{r}
plot(brmCatMod)
```

### Linear regression with two independent categorical variables - with `rethinking`

The `map()` function, standing for *Maximum a posteriori* fitting, in the first edition, has now been renamed to `quad()`, although the original function can still be called. If you are using an older version of `rethinking`, remember that just as for all the functions that has any potential to create ambiguity, call the function with the package name specified (e.g. `rethinking::map()`). Here I am using `quad()`.

The function renaming here is explained as in the [README for the `rethinking` package](https://github.com/rmcelreath/rethinking/blob/master/README.md):

> It was renamed, because the name map was misleading. This function produces quadratic approximations of the posterior distribution, not just maximum a posteriori (MAP) estimates.

Consider model:

$$
y_i \sim \text{Normal}(\mu_i, \sigma) \\
\mu_i = \alpha + \beta_a x_{a,i} + \beta_b x_{b,i} + \beta_{ab} x_{a,i} + x_{a,i} \\
\alpha \sim \text{Normal}(0, 5) \\
\beta \sim \text{Normal}(0, 5)\\
\sigma \sim \text{Uniform}(0, 5)\\
$$

In `rethinking`, a prior of `a ~ dnorm(0, 10)` translates to `dnorm(a, mean = 0, sd = 10)`.

```{r}
library('rethinking')
fit5 <- quap(
    alist(
        effect ~ dnorm(mu, sigma), 
        mu <- a + ba * drugA + bb * drugB + bab * drugA * drugB, 
        a ~ dnorm(0, 5), 
        ba ~ dnorm(0, 5), bb ~ dnorm(0, 5), bab ~ dnorm(0, 5), 
        sigma ~ dunif(0, 5)), 
    data = alldat)
fit10 <- quap(
    alist(
        effect ~ dnorm(mu, sigma), 
        mu <- a + ba * drugA + bb * drugB + bab * drugA * drugB, 
        a ~ dnorm(0, 10), 
        ba ~ dnorm(0, 10), bb ~ dnorm(0, 10), bab ~ dnorm(0, 10), 
        sigma ~ dunif(0, 5)), 
    data = alldat)
```

### Side-by-side comparison of the fittings

```{r}
# stats::lm()
print(twovarMod$coefficients)
# brms::brm()
print(fixef(brmCatMod))
# rethinking::quap()
print(rethinking::precis(fit5))
print(rethinking::precis(fit10))
```

## Supplementary information

### Installation of package `brms`

The package `brms` is available on CRAN. Install by running `install.packages('brms')` and verify the installation by running `library('brms')`. 

## For future references

### The session info when the documentation is written

The current version is written on 20200526, with a recent `R 4.0` release.

Before starting the installation process, out of personal maintenance preferences, I removed `R` (then version 3.6), RStudio, and `Rtools` and installed the latest versions of `R` (now version 4.0) and RStudio (now version 1.3.959). For `Rtools`, notice the version compatibility with `R 4.0`. `Rtools40` installation is described above.

As a reference point for the environment-specific fixes above, I am recording my 20200526 session information for the "this works on my machine" setting, to be compared to *your* session information if needed.

My 20200526 session info:

``` {r, eval = FALSE}
> sessionInfo()
R version 4.0.0 (2020-04-24)
Platform: x86_64-w64-mingw32/x64 (64-bit)
Running under: Windows 10 x64 (build 18362)

Matrix products: default

locale:
[1] LC_COLLATE=English_United States.1252  LC_CTYPE=English_United States.1252   
[3] LC_MONETARY=English_United States.1252 LC_NUMERIC=C                          
[5] LC_TIME=English_United States.1252    

attached base packages:
[1] parallel  stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
[1] rethinking_2.01    dagitty_0.2-2      rstan_2.19.3       ggplot2_3.3.0     
[5] StanHeaders_2.19.2 devtools_2.3.0     usethis_1.6.1     

loaded via a namespace (and not attached):
 [1] Rcpp_1.0.4.6       mvtnorm_1.1-0      lattice_0.20-41    prettyunits_1.1.1 
 [5] ps_1.3.3           assertthat_0.2.1   rprojroot_1.3-2    digest_0.6.25     
 [9] packrat_0.5.0      V8_3.0.2           R6_2.4.1           backports_1.1.7   
[13] stats4_4.0.0       evaluate_0.14      coda_0.19-3        pillar_1.4.4      
[17] rlang_0.4.6        curl_4.3           rstudioapi_0.11    callr_3.4.3       
[21] rmarkdown_2.1      desc_1.2.0         loo_2.2.0          munsell_0.5.0     
[25] compiler_4.0.0     xfun_0.14          pkgconfig_2.0.3    pkgbuild_1.0.8    
[29] shape_1.4.4        htmltools_0.4.0    tidyselect_1.1.0   tibble_3.0.1      
[33] gridExtra_2.3      matrixStats_0.56.0 fansi_0.4.1        crayon_1.3.4      
[37] dplyr_0.8.5        withr_2.2.0        MASS_7.3-51.5      grid_4.0.0        
[41] jsonlite_1.6.1     gtable_0.3.0       lifecycle_0.2.0    magrittr_1.5      
[45] scales_1.1.1       cli_2.0.2          fs_1.4.1           remotes_2.1.1     
[49] testthat_2.3.2     ellipsis_0.3.1     vctrs_0.3.0        boot_1.3-24       
[53] tools_4.0.0        glue_1.4.1         purrr_0.3.4        rsconnect_0.8.16  
[57] processx_3.4.2     pkgload_1.0.2      yaml_2.2.1         inline_0.3.15     
[61] colorspace_1.4-1   sessioninfo_1.1.1  memoise_1.1.0      knitr_1.28
```

Your current session info:

``` {r}
sessionInfo()
```

### Do-dos:

1. [ ] add resource links
2. [ ] add support to other operation systems
3. [ ] add repo-level "getting started with R and RStudio" and "Some struggles with Windows and how to solve hopefully under 5 hours"
4. [ ] double check encoding and line breaks since this is written on my Win10 machine
5. [ ] add "slides" for chapter 1 (and organize: as a separate file?)
6. [ ] add pointers to lab meeting slides on regression (20200518) and BRMS (20190722)
7. [ ] add model selection examples in BRMS?
8. [ ] add references to detailed contents as the contents build up for the following chapters
